---
title: "Machine Learning: Classifying Weightlifting Exercises"
author: "Don Mowbray"
output: html_document
---

<center><h1>Classifying Weightlifting Exercises</h1></center>
<center><h4>Practical Machine Learning, Coursera</h4></center>
<center><h4>Don Mowbray</h4></center>
<center><h4>March 21, 2015</h4></center>

## Overview ##

This report examines the Weight Lifting Exercises (WLE) Dataset (
[citation](http://groupware.les.inf.puc-rio.br/har) ) to predict
weightlifting activity quality from several activity
monitors. Activity quality is represented by the variable *classe*, a
factor with five levels representing how a barbell lifting activity
was conducted:

* according to the specification (Class A), 
* elbows to the front (Class B), 
* lifting halfway (Class C), 
* lowering halfway (Class D)
* hips to the front (Class E).


## Loading and Cleaning the DataSet ##

The following code was used to load and clean the datasets. Cleaning
the dataset involved removing non-predictive columns (user name,
timestamp, etc.) and deleting columns with missing (NA) values.

```{r loadLibraries, hide=TRUE}
# load the libraries required for this project
library(AppliedPredictiveModeling)
library(caret)
library(rpart)
library(rattle)
```

```{r LoadAndCleanData}
# load the csv training and testing sets
training <- read.csv('~/Downloads/pml-training.csv', na.strings=c("NA", ""), header=TRUE)
testing  <- read.csv('~/Downloads/pml-testing.csv', na.strings=c("NA", ""), header=TRUE)

# remove non-predictive columns
training <- training[,-c(1:7)]
testing  <- testing[,-c(1:7)]  

# Delete columns with missing values
training <-training[,colSums(is.na(training)) == 0]
testing <-testing[,colSums(is.na(testing)) == 0]
```


## Partitioning for Cross Validation ##

*createDataPartition* is used to create a balanced 75/25 split of the
training dataset into training and testing subsamples. Balanced
implies that random sampling occurs within each class to preserve the
overall class distribution of the data. The resulting training subset
will be used to construct our models, and the testing subset will be
employed for cross validation to estimate the accuracy of our
candidate models with data that is independent from the training
subset used to fit our models. Thus, cross validation allows us to
select the most accurate model to apply to the final testing dataset.

```{r CreateDataPartition}
set.seed(1234)
inTrain = createDataPartition(training$classe, p = 0.75, list=FALSE)
subTraining = training[inTrain,]
subTesting = training[-inTrain,]
```


## Classification Tree Model ###

First we fit a classification tree against the training subset.

```{r TreeFit}
treeModel <- rpart(classe ~ ., data=subTraining, method="class")
print(treeModel, digits=3)
```

Next we formulate predictions from our testing subset. Then we we use
cross validation to estimate the tree model's accuracy and out of
sample error.

```{r TreePredict}
treePredict <- predict(treeModel, subTesting, type = "class")
cmTree <- confusionMatrix(treePredict, subTesting$classe)
cmTree
```

The resulting confusion matrix suggests this model has an accuracy of
`r round(cmTree$overall['Accuracy'], digits=3)`. Calculating the out
of sample error as 1 - accuracy, this yields a out of sample error of
`r round(100*(1.0 - cmTree$overall['Accuracy']), digits=3)` percent.

## Random Forest Model ##

Next we fit a random forest against the training subset. During
preprocessing, principal component analysis (PCA) is used to reduce
the dimensionality of the data (feature reduction) while preserving
the data's essential variance. With the *cv* method parameter, we tune
the training function to use k-folds cross validation. We choose an
*ntree* parameter of 50 to balance model accuracy with computational
performance.

```{r RandomForestFit}
tc <- trainControl(method="cv", number=5, verboseIter=FALSE , preProcOptions="pca")
rfModel <- train(classe ~ ., data=subTraining, method="rf", trControl=tc, ntree=50)
varImp(rfModel$finalModel, top=20) # show variable importance
```

Next, we run this model against our testing subset to estimate its
accuracy.

```{r RandomForestPredict}
rfPredict <- predict(rfModel, subTesting);
cmRF <- confusionMatrix(rfPredict, subTesting$classe)
cmRF
```

Based on the above, we estimate this model has an accuracy of 
`r round(cmRF$overall['Accuracy'], digits=3)` and an out of sample error of 1 - accuracy =
`r round(100.0*(1.0 - cmRF$overall['Accuracy']), digits=3)` percent.

## Test Case Model Performance ##

Since the random forest model proved to be the most accurate, we
select that model and use it to make predictions against the final
test dataset.

```{r testCases}
rfTestPredict <- predict(rfModel, testing);
rfTestPredict
```

The following code is used to submit our prediction results:

```{r submitResults}
for(i in 1:length(rfTestPredict)) {
   filename <- sprintf( "PMLtest%d", i);
   write.table(rfTestPredict[i], filename, append=FALSE, quote=FALSE, row.names=FALSE, col.names=FALSE);
}
```

The random forest model predicted all 20 of the 20 test cases correctly.

## Citations ##

Velloso, E.; Bulling, A.; Gellersen, H.; Ugulino, W.; Fuks,
H. Qualitative Activity Recognition of Weight Lifting
Exercises. Proceedings of 4th International Conference in Cooperation
with SIGCHI (Augmented Human '13) . Stuttgart, Germany: ACM SIGCHI,
2013.